Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2022-04-19 doc/manuscript/ESSDmanuscript.Rmd
describing land classification

## Land use categories
IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land). 
Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ* [@ipcc_good_2003].
Definitions of forest are allowed to vary by country, but must be applied consistently.
Forest Land includes land where vegetation temporarily falls below the threshold values for forest (e.g., due to disturbance), but is expected to exceed those thresholds in the future [@ipcc_good_2003].

Inventories are conducted only for managed land, which is defined by IPCC to include "all forests subject to some kind of human interactions (notably commercial management, harvest of industrial round-wood (logs) and fuelwood, production and use of wood commodities, and forest managed for amenity value or environmental protection if specified by the country), with defined geographical boundaries" [@ipcc_good_2003].
As such, managed land constitutes all or the majority of Forest Land in most countries; however, many countries have not yet reported their approach for defining managed land [@ogle_delineating_2018; @deng_comparing_2021].

@@

For the majority of fields, contents of the field in ForC was transferred directly into an EFDB field, either as the only contents of that field or as part of a composite record.
For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field `Region/Regional conditions` (`r table_ForCfieldmapping`).
In cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present information required to calculate these (standard error or n and standard deviation), we calculated the 95% confidence intervals and populated the EFDB field with this information (noting the calculation in the EFDB field `Comments from Data Provider`).
For some fields, simple conditional logic was used to populate EFDB fields based on ForC records. 
For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (`Gases` field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~. 
There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (section `r section_C_variable_mapping`) and (2) determination of land use categories (section `r section_land_use_mapping`).

@@

### Land use categories

@@

The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [*IPCC 2006 full report REF*]. 
This definition is applied differently across countries, and is not clearly defined by the majority of governments [@ogle_delineating_2018]. 
Given this, and because the IPCC definition of management does not necessarily match that which would be reported in scientific publications and hence in ForC, we do not transfer any classification of land management status from ForC to the EFDB, but do provide auxiliary info that may be useful in making this determination (e.g., geographical location).

>>>>

## Land classification
IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land). 
Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ*, and shorter-stature natural vegetation would be classified as Grassland [@ipcc_good_2003].
Definitions of forest are allowed to vary by country, but must be applied consistently.
Forest Land includes land where vegetation temporarily falls below the threshold values for forest (e.g., due to disturbance), but is expected to exceed those thresholds in the future [@ipcc_good_2003].

The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [@ipcc_2006_2006]. 
This expansive definition of managed land implies that the majority of Forest Land in most countries is managed [e.g.,].
However, the definition is applied differently across countries, and the majority of governments have yet to report their approach for defining managed land or provide maps of managed land [@ogle_delineating_2018; @deng_comparing_2021].

@@

For the majority of fields, contents of the field in ForC was transferred directly into an EFDB field, either as the only contents of that field or as part of a composite record.
For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field `Region/Regional conditions` (`r table_ForCfieldmapping`).
In cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present information required to calculate these (standard error or n and standard deviation), we calculated the 95% confidence intervals and populated the EFDB field with this information (noting the calculation in the EFDB field `Comments from Data Provider`).
For some fields, simple conditional logic was used to populate EFDB fields based on ForC records. 
For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (`Gases` field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~. 
There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (section `r section_C_variable_mapping`) and (2) land classification (section `r section_land_use_mapping`).

@@

### Land classification

Determination of the IPCC land-use category (i.e., Forest Land, Grassland, Wetlands, Cropland, Settlements, or Other Land; section `r  section_land_use_categories`) was made based on the categorical ForC field *dominant.life.form*, sometimes drawing upon stand age.
Records with "woody" *dominant.life.form* were classified as Forest Land.
Those with *dominant.life.form* of "woody+grass", which in ForC is indicative of anything from a shrub-encroached grassland to a tree-dominated savanna, were given dual classification of Forest Land and Grassland.
This dual classification indicates that records may be relevant to either category depending on the definition of forest applied (varies by country).
For (rare) cases where *dominant.life.form* was grass and stand age was greater than zero, indicative of early successional vegetation, we assigned a classification of Forest Land, consistent with the IPCC definition that Forest Land includes land expected to succeed to forest.
Cases where *dominant.life.form* was grass or crop and stand age was zero were indicative of a control for studies of forest regrowth following agricultural abandonment, and were classified as Grassland and Cropland, respectively.

*(now describe sub-category classification, which depends on past land use)*

@@

Given the lack of public information needed to determine whether lands are classified as mangaged [@ogle_delineating_2018; @deng_comparing_2021], and because the IPCC's definition of managed land is more expansive than is commonly applied in the scientific literature and hence in ForC, we did not transfer any classification of land management status from ForC to the EFDB.
However, we do provide auxiliary information that should be useful in making this determination, including geographical location and notable disturbance events.

### 2022-04-19 doc/manuscript/ESSDmanuscript.tex
describing land classification

\subsection{Land use categories}

IPCC defines land-use categories to include six categories -- Forest
Land, Grassland, Wetlands, Cropland, Settlements, and Other Land
\citep{ipcc_2006_2006}. Sub-divisions include land that has remained in
a particular category for \textgreater20 years (e.g., Forest Land
remaining Forest Land) and land that has been converted from one
category to another in the past 20 years (e.g., Cropland converted to
Forest Land). Forest Land is defined as at least 10-30\% crown cover of
trees with potential to reach a minimum height of 2-5 m \emph{in situ}
\citep{ipcc_good_2003}. Definitions of forest are allowed to vary by
country, but must be applied consistently. Forest Land includes land
where vegetation temporarily falls below the threshold values for forest
(e.g., due to disturbance), but is expected to exceed those thresholds
in the future \citep{ipcc_good_2003}.

Inventories are conducted only for managed land, which is defined by
IPCC to include ``all forests subject to some kind of human interactions
(notably commercial management, harvest of industrial round-wood (logs)
and fuelwood, production and use of wood commodities, and forest managed
for amenity value or environmental protection if specified by the
country), with defined geographical boundaries'' \citep{ipcc_good_2003}.
As such, managed land constitutes all or the majority of Forest Land in
most countries; however, many countries have not yet reported their
approach for defining managed land
\citep{ogle_delineating_2018, deng_comparing_2021}.

@@

To transfer complete, reviewed ForC records into EFDB, we created R
scripts to restructure ForC records and populate EFDB's bulk import form
(``EFDB bulk import.xlsx''). Criteria for data transfer were that (1)
records had been checked against the original study and determined to be
complete and correct (\texttt{EFDB.ready} field in
\texttt{ForC\_citations} table), (2) the original study presented values
in tables or text, as opposed to the values having been digitized from
graphs or calculated based on related variables
(\texttt{data.location.within.source} field in
\texttt{ForC\_measurements} table), and (3) the records had not
previously been sent to EFDB. Once converted into EFDB format, the
records were reviewed and then sent to the IPCC's Technical Support Unit
for inclusion in EFDB. Complete records needed to be reviewed by the
EFDB editorial board prior to posting in the database -- a process that
lags behind records transfer and had not yet been completed for all
records sent as of April 18, 2022.

@@

For the majority of fields, contents of the field in ForC was
transferred directly into an EFDB field, either as the only contents of
that field or as part of a composite record. For example, ten ForC
fields describing site location, climate, and edaphic properties all
mapped into the EFDB field \texttt{Region/Regional\ conditions}
(Appendix B). In cases where original studies did not present 95\%
confidence intervals (required by IPCC when available) but did present
information required to calculate these (standard error or n and
standard deviation), we calculated the 95\% confidence intervals and
populated the EFDB field with this information (noting the calculation
in the EFDB field \texttt{Comments\ from\ Data\ Provider}). For some
fields, simple conditional logic was used to populate EFDB fields based
on ForC records. For example, for stock variables presented in the
original publication in units of dry organic matter mass (as opposed to
C), several greenhouse gasses (CO\textsubscript{2}, CO,
CH\textsubscript{4}, NO, NO\textsubscript{2}, N\textsubscript{2}O) were
entered in the EFDB field indicating the greenhouse gases to which the
record could be pertinent (\texttt{Gases} field) because these values
could be used in calculations of greenhouse gas emissions from biomass
burning \citep{ipcc_2006_2006}; otherwise, the only pertinent greenhouse
gas would be CO\textsubscript{2}. There were two cases in which more
complex mapping was required: (1) mapping of C cycle variables (section
4.1.1) and (2) determination of land use categories (section 4.1.2).

@@

\subsubsection{Land use categories}

@@

The UNFCCC requires greenhouse gas reporting for all managed lands in a
country, where management is defined as ``human interventions and
practices have been applied to perform production, ecological or social
functions'' {[}\emph{IPCC 2006 full report REF}{]}. This definition is
applied differently across countries, and is not clearly defined by the
majority of governments \citep{ogle_delineating_2018}. Given this, and
because the IPCC definition of management does not necessarily match
that which would be reported in scientific publications and hence in
ForC, we do not transfer any classification of land management status
from ForC to the EFDB, but do provide auxiliary info that may be useful
in making this determination (e.g., geographical location).

@@

As of April 18, 2022, ForC (v4.0) contained NA independent records
(39838 total), NA of which were for variables relevant to EFDB (Fig. 1).
These records were distributed across all forested continents and
ecozones, with particularly high concentrations in
\emph{{[}ecozone(s){]}} and low concentrations in
\emph{{[}ecozone(s){]}} (Fig. 2). The most widely represented forest
type was \emph{{[}type{]}}, followed by \emph{{[}type{]}} and
\emph{{[}type{]}} (Fig. 3).

@@

As of April 18, 2022, we had reviewed and sent NA records to EFDB, 73 of
which have been reviewed, accepted, and posted (Figs. 2-3, Table 2).
\emph{{[}DETAILS{]}}

@@

\caption{\label{tab:table_recommendations}\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance. } ....}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\hline
\textbf{criteria} & \textbf{recommendation} & \textbf{rationale}\\
\hline
variables to include & When possible, calculate and present all relevant variables that can be readily caluclated based on available data. & Estimates of relevant variables are not always calcualted.\\
\hline
forest census methods & Census stems down to the smallest diameter feasible. & IPCC biomass pool definition includes all living vegetation\\
\hline
 & taxa to include: & IPCC biomass pool definition includes all living vegetation\\
\hline
dead organic matter sampling & include damage estimates on standing dead trees & \\
\hline
 & adopt IPCC recommendations for min diameter of deadwood & \\
\hline
belowground sampling & Select soil sampling increments to include a cutoff at 30 cm depth. & \\
\hline
reporting variiables & Present each variable individually, as opposed to requiring that variables of interest be calculated from related variables. & EFDB requires that values in the database be presented in the original article, and cannot accept subsequent calculations.\\
\hline
reporting estimates & Report all relevant values in tables, text, or supplementary tables/ data files, as opposed to in figures only. & EFDB does not acceptvalues digitized from figures.\\
\hline
reporting confidence intervals & Report 95\% confidence intervals, standard error, or standard deviation and sample size. & EFDB requires confidence invervals, when possible.\\
\hline
\end{tabu}
\end{table}

>>>>

\subsection{Land classification}

IPCC defines land-use categories to include six categories -- Forest
Land, Grassland, Wetlands, Cropland, Settlements, and Other Land
\citep{ipcc_2006_2006}. Sub-divisions include land that has remained in
a particular category for \textgreater20 years (e.g., Forest Land
remaining Forest Land) and land that has been converted from one
category to another in the past 20 years (e.g., Cropland converted to
Forest Land). Forest Land is defined as at least 10-30\% crown cover of
trees with potential to reach a minimum height of 2-5 m \emph{in situ},
and shorter-stature natural vegetation would be classified as Grassland
\citep{ipcc_good_2003}. Definitions of forest are allowed to vary by
country, but must be applied consistently. Forest Land includes land
where vegetation temporarily falls below the threshold values for forest
(e.g., due to disturbance), but is expected to exceed those thresholds
in the future \citep{ipcc_good_2003}.

The UNFCCC requires greenhouse gas reporting for all managed lands in a
country, where management is defined as ``human interventions and
practices have been applied to perform production, ecological or social
functions'' \citep{ipcc_2006_2006}. This expansive definition of managed
land implies that the majority of Forest Land in most countries is
managed {[}e.g.,{]}. However, the definition is applied differently
across countries, and the majority of governments have yet to report
their approach for defining managed land or provide maps of managed land
\citep{ogle_delineating_2018, deng_comparing_2021}.

@@

To transfer complete, reviewed ForC records into EFDB, we created R
scripts to restructure ForC records and populate EFDB's bulk import form
(``EFDB bulk import.xlsx''). Criteria for data transfer were that (1)
records had been checked against the original study and determined to be
complete and correct (\texttt{EFDB.ready} field in
\texttt{ForC\_citations} table), (2) the original study presented values
in tables or text, as opposed to the values having been digitized from
graphs or calculated based on related variables
(\texttt{data.location.within.source} field in
\texttt{ForC\_measurements} table), and (3) the records had not
previously been sent to EFDB. Once converted into EFDB format, the
records were reviewed and then sent to the IPCC's Technical Support Unit
for inclusion in EFDB. Complete records needed to be reviewed by the
EFDB editorial board prior to posting in the database -- a process that
lags behind records transfer and had not yet been completed for all
records sent as of April 19, 2022.

@@

For the majority of fields, contents of the field in ForC was
transferred directly into an EFDB field, either as the only contents of
that field or as part of a composite record. For example, ten ForC
fields describing site location, climate, and edaphic properties all
mapped into the EFDB field \texttt{Region/Regional\ conditions}
(Appendix B). In cases where original studies did not present 95\%
confidence intervals (required by IPCC when available) but did present
information required to calculate these (standard error or n and
standard deviation), we calculated the 95\% confidence intervals and
populated the EFDB field with this information (noting the calculation
in the EFDB field \texttt{Comments\ from\ Data\ Provider}). For some
fields, simple conditional logic was used to populate EFDB fields based
on ForC records. For example, for stock variables presented in the
original publication in units of dry organic matter mass (as opposed to
C), several greenhouse gasses (CO\textsubscript{2}, CO,
CH\textsubscript{4}, NO, NO\textsubscript{2}, N\textsubscript{2}O) were
entered in the EFDB field indicating the greenhouse gases to which the
record could be pertinent (\texttt{Gases} field) because these values
could be used in calculations of greenhouse gas emissions from biomass
burning \citep{ipcc_2006_2006}; otherwise, the only pertinent greenhouse
gas would be CO\textsubscript{2}. There were two cases in which more
complex mapping was required: (1) mapping of C cycle variables (section
4.1.1) and (2) land classification (section 4.1.2).

@@

\subsubsection{Land classification}

Determination of the IPCC land-use category (i.e., Forest Land,
Grassland, Wetlands, Cropland, Settlements, or Other Land; section 2.2)
was made based on the categorical ForC field \emph{dominant.life.form},
sometimes drawing upon stand age. Records with ``woody''
\emph{dominant.life.form} were classified as Forest Land. Those with
\emph{dominant.life.form} of ``woody+grass'', which in ForC is
indicative of anything from a shrub-encroached grassland to a
tree-dominated savanna, were given dual classification of Forest Land
and Grassland. This dual classification indicates that records may be
relevant to either category depending on the definition of forest
applied (varies by country). For (rare) cases where
\emph{dominant.life.form} was grass and stand age was greater than zero,
indicative of early successional vegetation, we assigned a
classification of Forest Land, consistent with the IPCC definition that
Forest Land includes land expected to succeed to forest. Cases where
\emph{dominant.life.form} was grass or crop and stand age was zero were
indicative of a control for studies of forest regrowth following
agricultural abandonment, and were classified as Grassland and Cropland,
respectively.

\emph{(now describe sub-category classification, which depends on past
land use)}

@@

Given the lack of public information needed to determine whether lands
are classified as mangaged
\citep{ogle_delineating_2018, deng_comparing_2021}, and because the
IPCC's definition of managed land is more expansive than is commonly
applied in the scientific literature and hence in ForC, we did not
transfer any classification of land management status from ForC to the
EFDB. However, we do provide auxiliary information that should be useful
in making this determination, including geographical location and
notable disturbance events.

@@

As of April 19, 2022, ForC (v4.0) contained NA independent records
(39838 total), NA of which were for variables relevant to EFDB (Fig. 1).
These records were distributed across all forested continents and
ecozones, with particularly high concentrations in
\emph{{[}ecozone(s){]}} and low concentrations in
\emph{{[}ecozone(s){]}} (Fig. 2). The most widely represented forest
type was \emph{{[}type{]}}, followed by \emph{{[}type{]}} and
\emph{{[}type{]}} (Fig. 3).

@@

As of April 19, 2022, we had reviewed and sent NA records to EFDB, 73 of
which have been reviewed, accepted, and posted (Figs. 2-3, Table 2).
\emph{{[}DETAILS{]}}

@@

\caption{\label{tab:table_recommendations}\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance. } ....}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\hline
\textbf{criteria} & \textbf{recommendation} & \textbf{rationale}\\
\hline
variables to include & When possible, calculate and present all relevant variables that can be readily caluclated based on available data. & Estimates of relevant variables are not always calcualted.\\
\hline
forest census methods & Adopt IPCC guidelines (country-specific) for minimum stem size in censues in census and reporting. Ideally, census stem down to the smallest diameter feasible. & IPCC biomass pool definition includes all living vegetation\\
\hline
 & *taxa to include* & \\
\hline
dead organic matter sampling & *include damage estimates on standing dead trees* & \\
\hline
 & Adopt IPCC recommendations for minimum diameter of deadwood (country-specific, default 10 cm). & \\
\hline
belowground sampling & Select and report soil sampling increments to include a cutoff at 30 cm depth (or country-specific depth). & IPCC biomass pool definition includes all living vegetation\\
\hline
reporting variiables & Present each variable individually, as opposed to requiring that variables of interest be calculated from related variables. & EFDB requires that values in the database be presented in the original article, and cannot accept subsequent calculations.\\
\hline
reporting estimates & Report all relevant values in tables, text, or supplementary tables/ data files, as opposed to in figures only. & EFDB does not acceptvalues digitized from figures.\\
\hline
reporting confidence intervals & Report 95\% confidence intervals, standard error, or standard deviation and sample size. & EFDB requires confidence invervals, when possible.\\
\hline
\end{tabu}
\end{table}
